---
title: "Programming concepts"
author: "Chris Paciorek"
date: "2022-09-06"
format:
  pdf:
    documentclass: article
    margin-left: 30mm
    margin-right: 30mm
    toc: true
  html:
    theme: cosmo
    css: ../styles.css
    toc: true
    code-copy: true
    code-block-background: true
execute:
  freeze: auto
---

[PDF](./unit5-programming.pdf){.btn .btn-primary}


> **Note**
> This Unit will be posted in phases as I get the material ready.
> 2022-09-09: Currently only Sections 1-6 are available.


References:

-   Books on R listed on the syllabus: Adler, Chambers, Wickham
-   [R intro manual](http://cran.r-project.org/doc/manuals/R-intro.html)
    and [R language
    manual](http://cran.r-project.org/doc/manuals/R-lang.html) (R-lang),
    both on CRAN.
-   Murrell, Introduction to Data Technologies

(Optional) Videos

There are various videos from 2020 in the bCourses Media Gallery that you
can use for reference if you want to. Note that I've reorganized the material in this Unit relative to 2020, so the section numbers and ordering in the videos may differ from that in the current Unit, but you should be able to match things up fairly easily.

1. Strings and regular expressions
3. Type/class coercion
4. Object-oriented programming - S3 classes
5. Object-oriented programming - R6 classes


# Overview

This unit covers a variety of programming concepts, illustrated in the
context of R and with comments about and connections to other languages. It also serves as a way to teach some advanced features of
R. In general the concepts are relevant in other languages, though other
languages may implement things differently. One of my goals for the unit is for us
to think about why things are the way they are in R. I.e., what
principles were used in creating the language and what choices were
made? While other languages use different principles and made different 
choices, understanding what one language does in detail will be helpful when you
are learning another language or choosing a language for a project.

I'll likely refer to R syntax as *statements* or *expressions*, meaning
 any code that is a valid, complete R expression. But note that the term *expression* also means a specific type of object
within the R language, as seen late in this Unit when we discuss computing on the language.

# 1. Text manipulation, string processing and regular expressions (regex)

Text manipulations in R have a number of things in common with Python,
Perl, and UNIX, as many of these evolved from UNIX. When I use the term
*string* here, I'll be referring to any sequence of characters that may
include numbers, white space, and special characters, rather than to the
character class of R objects. The string or strings will generally be
stored as an R character vector.

## String processing and regular expressions in R

For details of string processing in R, including use of regular expressions, see the [string
processing tutorial](https://berkeley-scf.github.io/tutorial-string-processing).
(You can ignore the sections on Python if you wish.) That tutorial then refers to
the [bash shell tutorial](https://berkeley-scf.github.io/tutorial-using-bash/regex) 
for details on regular expressions.

In class we'll work through some problems in the string processing tutorial, focusing in
particular on the use of regular expressions with the *stringr* package.
This will augment our consideration of regular expressions in the shell, in particular
by seeing how we can replace patterns in addition to finding them. 

## Regex/string processing challenges

We'll work on these challenges (and perhaps one or two others) in class in the process of working
through the string processing tutorial.

1.  What regex would I use to find any number with or without a decimal place.

2.  Suppose a text string has dates in the form "Aug-3", "May-9", etc.
    and I want them in the form "3 Aug", "9 May", etc. How would I do
    this search and replace operation? (Alternatively, how could I do
    this without using regular expressions at all?)

## Side notes on special characters in R 

Recall that when characters are used for special purposes, we need to
'escape' them if we want them interpreted as the actual character. In what
follows, I show this in R, but similar manipulations are sometimes
needed in the shell and in Python.

This can get particularly confusing in R as the backslash is also used
to input special characters such as newline (`\n`) or tab (`\t`).

Here are some examples of using special characters.

> **Note**
> It is hard to compile the Rmd file correctly for these R chunks, so I am just pasting in the output from running in R 'manually' in some cases.)

```{r}
tmp <- "Harry said, \"Hi\""
## cat(tmp)   # prints out without a newline -- this is hard to show in the pdf
tmp <- "Harry said, \"Hi\".\n"
cat(tmp)      # prints out with the newline

tmp <- c("azar", "foo", "hello\tthere\n")
cat(tmp)
print(tmp)
grep("[\tz]", tmp)   ## search for a tab or a 'z'
```

As a result, in R we often need two backslashes when working with
regular expressions. In the next examples, the first backslash says to
interpret the next backslash literally, with the second backslash being
used to indicate that the caret (\^) should be interpreted literally and
not as a special character in the regular expression syntax.

```{r}
## Search for characters that are not 'z'
## (using ^ as regular expression syntax)
grep("[^z]", c("a^2", "93", "zzz", "zit", "azar"))

## Search for either a '^' (as a regular charcter) or a 'z':
grep("[\\^z]", c("a^2", "93", "zzz", "zit", "azar"))

## This fails (and the Rmd won't compile) because
## '\^' is not an escape sequence (i.e., a special character):
## grep("[\^z]", c("a^2", "93", "zit", "azar", "zzz"))
## Error: '\^' is an unrecognized escape in character string starting ""[\^"

## Search for exactly three characters
## (using . as regular expression syntax)
grep("^.{3}$", c("abc", "1234", "def"))

## Search for a period (as a regular character)
grep("\\.", c("3.9", "27", "4.2"))

## This fails (and the Rmd won't compile) because 
## '\.' is not an escape sequence (i.e., a special character):
## grep("\.", c("3.9", "27")))
## Error: '\.' is an unrecognized escape in character string starting ""\."
```

> **Challenge**
> Explain why we use a single backslash to get a newline and double backslash to write out a Windows path in the examples here:

```{r}
## Suppose we want to use a \ in our string:
cat("hello\nagain")
cat("hello\\nagain")

cat("My Windows path is: C:\\Users\\My Documents.")
```

For more information, see `?Quotes` in R and the subsections of the
string processing tutorial that discuss backslashes and escaping.

Advanced note: Searching for an actual backslash gets even more
complicated, because we need to pass two backslashes as the regular
expression, so that a literal backslash is searched for. However, to
pass two backslashes, we need to escape each of them with a backslash so
R doesn't treat each backslash as part of a special character. So that's
four backslashes to search for a single backslash! Yikes. One rule of
thumb is just to keep entering backslashes until things work!

```{r}
## Search for an actual backslash
tmp <- "something \\ other\n"
cat(tmp)

grep("\\\\", tmp)
try(grep("\\", tmp))
```

> **Warning**
> Be careful when cutting and pasting from documents that are not text
> files as you may paste in something that looks like a single or double
> quote, but which R cannot interpret as a quote because it's some other
> ASCII quote character. If you paste in a " from PDF, it will not be
> interpreted as a standard R double quote mark.

Similar things come up in the shell and in Python, but in the shell you
often don't need two backslashes. E.g. you could do this to look for a
literal \^ character.

```{bash, eval=FALSE}
grep '\^' file.txt
```

# 2. Interacting with the operating system and external code and configuring R 

## Interacting with the operating system

Scripting languages allow one to interact with the operating system in various ways.
Most allow you to call out to the shell to run arbitrary shell code and save results within your session.

I'll assume everyone knows about the following functions/functionality for interacting with the filesystem and file
in R: *getwd*, *setwd*, *source*, *pdf*, *save*, *save.image*, *load*.

-   To run UNIX commands from within R, use `system()`, as follows,
    noting that we can save the result of a system call to an R object:

    ```{r}
    system("ls -al")   ## results apparently not shown when compiled...
    files <- system("ls", intern = TRUE)
    files[1:5]
    ```

-   There are also a bunch of functions that will do specific queries of
    the filesystem, including

    ```{r}
    file.exists("unit2-dataTech.Rmd")
    list.files("../data")
    ```

-   There are some tools for dealing with differences between operating
    systems. *file.path* is a nice example:

    ```{r}
    list.files(file.path("..", "data"))
    ```

    It's best if you can to write your code in a way that is *agnostic* to the underlying operating system.

-   To get some info on the system you're running on:

    ```{r}
    Sys.info()
    ```

## Controlling the behavior of R

Scripting languages generally allow you to control/customize their behavior
in various ways by setting options.

-   To see some of the options that control how R behaves, try the
    *options* function. The *width* option changes the number of
    characters of width printed to the screen, while *max.print*
    revents too much of a large object from being printed to the
    screen. 

    ```{r}
    ## options()  # this would print out a long list of options
    options()[1:4]
    options()[c('width', 'digits')]

    ## Often it's nice to have more characters in each line on the screen,
    ## but that would cause overly lines in the compiled file.
    ## options(width = 120)

    options(max.print = 5000)
    ```

    The *digits* option changes the number of digits of numbers
    printed to the screen (but be careful as this can be deceptive if
    you then try to compare two numbers based on what you see on the
    screen).
    ```{r}
    options(digits = 3)
    a <- 0.123456; b <- 0.1234561
    a; b; a == b
    ```

    More on how to (and how not to) compare real-valued numbers on a computer in Unit 8.

-   Use `Ctrl-C` to interrupt execution. This will generally back out
    gracefully, returning you to a state as if the command had not been
    started. Note that if R is exceeding the amount of memory available, there can
    be a long delay. This can be frustrating, particularly since a
    primary reason you would want to interrupt is when R runs out of
    memory.

-   *sessionInfo* gives information on the current R session and can
    be very helpful for recording the state of your session (including package versions) to allow for reproducibility.

    ```{r}
    sessionInfo()
    ```

-   Any code that you wanted executed automatically when starting R can
    be placed in `~/.Rprofile` (or in individual, project-specific `.Rprofile` files in
    specific directories). This could include loading packages (see
    below), sourcing files that contain user-defined functions that you
    commonly use (you can also put the function code itself in
    `.Rprofile`), assigning variables, and specifying options via
    `options()`.

-   You can have an R script act as a shell script (like running a bash
    shell script) as follows. This will probably on work on Linux and
    Mac.

    1.  Write your R code in a text file, say `exampleRscript.R`.
    2.  As the first line of the file, include `#!/usr/bin/Rscript`
        (like `#!/bin/bash` in a bash shell file, as seen in Unit 2) or
        for more portability across machines, include
        `#!/usr/bin/env Rscript`.
    3.  Make the R code file executable with *chmod*:
        `chmod ugo+x exampleRscript.R`.
    4.  Run the script from the command line: `./exampleRscript.R`

    If you want to pass arguments into your script, you can do so as
    long as you set up the R code to interpret the incoming arguments:

    ```{r, eval=FALSE}
    args <- commandArgs(TRUE)

    ## Now args is a character vector containing the arguments.
    ## Suppose the first argument should be interpreted as a number 
    ## and the second as a character string and the third as a boolean:

    numericArg <- as.numeric(args[1])
    charArg <- args[2]
    logicalArg <- as.logical(args[3])

    cat("First arg is: ", numericArg, "; second is: ", charArg, 
        "; third is: ", logicalArg, ".\n")
    ```

    Now we can run it as follows in the shell:

    ```{bash}
    ./exampleRscript.R 53 blah T
    ./exampleRscript.R blah 22.5 t
    ```

## Interacting with external code

Scripting languages such as R, Python, and Julia allow you to call out to "external code",
which often means C or C++ (but also Fortran, Java and other languages). 

In fact, the predecessor language to R,
which was called 'S' was developed specifically (at AT&T's Bell Labs in the 1970s and 1980s) as an interactive
wrapper around Fortran, the numerical programming language most commonly used at the time (and still widely relied on today in various legacy codes). 

Calling out to external code is particularly important in languages like R and Python that are often much slower
than compiled code and less important in a fast language like Julia (which uses Just-In-Time compilation -- more on that later).

In R, one can call directly out to C or C++ code using *.Call* or one can use the [Rcpp package](https://adv-r.hadley.nz/rcpp.html). *Rcpp* is specifically designed to be able to write C++ code that feels somewhat like writing R code and where it is very easy to pass data between R and C++. 

In Python, one can [directly call out to C or C++ code](https://docs.python.org/3/extending/extending.html) or one can use *Cython* to interact with C. With Cython, one can:
  - Have Cython automatically translate Python code to C, if you provide type definitions for your variables.
  - Define C functions that can be called from your Python code.

# 3. Packages and namespaces

Scripting languages that become popular generally have an extensive collection
of add-on packages available online (the causal relationship of the popularity and
the extensive add-on packages goes in both directions). Packages need to
be *installed* (once) on your computer and *loaded* (every time you start a new session).

A big part of R's popularity is indeed the extensive collection of add-on
packages on [CRAN](https://cran.r-project.org) (and GitHub and elsewhere) that provide
much of R's functionality. To make use of a package it needs to be
installed on your system (using *install.packages* once only) and
loaded into R (using *library* every time you start R).

Some packages are *installed* by default with R and of these, some are
*loaded* by default, while others require a call to *library*. 

If you want to sound like an R expert, make sure to call them *packages*
and not *libraries*. A *library* is the location in the directory
structure where the packages are installed/stored.


## Loading packages

You can use *library* to either (1) make a package available (loading
it), (2) get an overview of the package, or (3) (if called without
arguments) to see all the installed packages.

```{r, eval=FALSE}
library(dplyr)              # load the package
library(help = dplyr)       # get some help info about the package
```

Packages in R (and in Python, Julia, etc.) may be installed in various places
on the filesystem, and it sometimes it is helpful (e.g., if you end up with multiple
versions of a package installed on your system) to be able to figure out 
where on the filesystem the package is being loaded from. 
If you run `library()`, you'll notice that some of the packages are in a
system directory and some are in your home directory. 

`.libPaths()` shows where R
looks for packages on your system and `searchpaths()` shows where
individual packages currently loaded in your session have been loaded from. The help information 
for *.libPaths* gives some information about how R decides what locations
to look in for packages (and how you can modify that).

```{r}
.libPaths()
searchpaths()
```

## Installing packages

If a package is on CRAN but not on your system, you can install it
easily (usually). You don't need root permission on a machine to install
a package (though sometimes you run into hassles if you are installing
it just as a user, so if you have administrative privileges it may help
to use them). Of course in RStudio, you can install via the GUI. 

Packages often depend on other packages. In general, if one package depends on another,
R will install the dependency automatically, but sometimes you'll need to 
install a dependency yourself.  In general, package dependencies
are handled very cleanly in R without you having having to worry much about it;
this is less the case in Python.


Note that R will generally install the package in a reasonable place by default
but you can control where it is installed using the *lib* argument.

```{r, eval=FALSE}
install.packages('dplyr', lib = '~/Rlibs') # ~/Rlibs needs to exist!
```

You can also download the zipped source file from CRAN and install from
the file; see the help page for *install.packages*. This is called
"installing from source". On Windows and Mac, you'll need to do
something like this:

```{r, eval=FALSE}
install.packages('dplyr_VERSION.tar.gz', repos = NULL, type = 'source')
```

This can be handy if you need to install [an older version of a package](https://cran.r-project.org/src/contrib/Archive)
for reproducibility or because of some dependency incompatibility.

If you've downloaded the binary package (files ending in .tgz for Mac
and .zip for Windows) and want to install the package directly from the
file, use the syntax above but omit the `type= 'source'` argument.

### Source vs. binary packages

The difference between a *source* package and a *binary* package is that
the source package has the raw R (and C and Fortran, in some cases) code
as text files while the binary package has all the code in a
binary/non-text format, including that any C and Fortran code will have already been
compiled. To install a source package with C or Fortran code in it,
you'll need to have developer/command-line tools (e.g., *XCode* on Mac
or *Rtools.exe* on Windows) installed on your system so that you have a
compiler.

## Managing packages using package managers

For reproducibility, it's important to know the versions of the packages you use (and the version of R).
Package managers make it easy to do this. Some useful packages that do package management in R are *checkpoint*, *renv*, and *packrat*. The basic commonality is that they try to make it easy to 'freeze' the versions of hte packages you are using, record that information, and restore the versions (potentially on some other machine and by some user other than yourself). The package manager may tell you where the packages are installed, but you can always verify things with `.libPaths()`.

In Python, you can set up and manage isolated environments in which you can control the package versions using virtualenvs or Conda environments.


## Package namespaces

The objects in a package (primarily functions, but also data) are in
their own workspaces, and are accessible after you load the package
using `library()`, but are not directly visible when you use `ls()`. In
other words, each package has its own *namespace*. Namespaces help
achieve modularity and avoid having zillions of objects all reside in
your workspace.  If we want to see the objects in a package's namespace, we
can do the following:

```{r}
search()
## ls(pos = 4) # for the stats package
ls(pos = 4)[1:5] # just show the first few
ls("package:stats")[1:5] # equivalent
ls("package:stats", pattern = "^lm")
```

### Why have namespaces?

We'll talk more about namespaces when we talk about variable scope and
environments. But as some motivation for why this is useful, consider the following.

The *lm* function calls the *lm.fit* function to calculate the least squares solution in regression.

Suppose we write our own *lm.fit* function that does something else:

```{r}
lm.fit <- function(x)
    print('hi')

x <- 7
lm.fit(x)
```

One might expect that if one now uses `lm()` to fit a regression, that it wouldn't work correctly because we have an *lm.fit* function in our workspace that doesn't calculate the least squares solution. But it works just fine (see below), because *lm* and *lm.fit* are in the *stats* package namespace (see above) and R's scoping rules (more later) ensure that the *lm.fit* that is found when I run *lm* is the *lm.fit* needed to run the regression and not my silly *lm.fit* function in current workspace.

```{r}
n <- 10
x <- runif(n)
y <- runif(n)
mod <- lm(y ~ x)
mod
```

### Namespace resolution

Standard practice in R has generally been to load a package and then use any of the items in the package namespace directly, e.g.,

```{r}
library(stringr)
str_detect("hello there", "hello")
```

However, particularly if you're using the package in only a limited way, it can be a nice idea to not load the entire package and instead use the namespace resolution operator in a style that might remind you of Python and some other languages:

```{r}
stringr::str_detect("hello there", "hello")
```

```{python}
import numpy as np
x = np.ndarray([0,3,5])
```

Of course in Python you could also load the entire package (i.e., import the entire namespace), though it's not standard practice:

```{python}
from numpy import *
## OR: from numpy import ndarray
x = ndarray([0,3,5])
```

Loading entire packages often causes 'name collisions' where there are multiple functions (or variables, more genreally) that have the same name. This can be confusing. We'll see how R determines what function to use later in the Unit.

# 4. Types and data structures

## Data structures

Please see the [data structures section of Unit 2](unit2-dataTech.html#data-structures) for some general discussion of data structures.

We'll also see more complicated data structures when we consider objects in the next section on object-oriented programming.


## Types and classes

### Overview and static vs. dynamic typing

The term 'type' refers to how a given piece of information is stored and what operations can be done with the information.
'Primitive' types are the most basic types that often relate directly to how data are stored in memory or on disk (e.g., boolean, integer, numeric (real-valued, aka *double* or *floating point*), character, pointer (aka *address*, *reference*).

In compiled languages like C and C++, one has to define the type of each variable. Such languages are *statically* typed.
Interpreted (or scripting) languages such as Python and R have *dynamic* types. One can associate different types of information with a given variable name at different times and without declaring the type of the variable:

```{r}
x <- 'hello'
print(x)
x <- 7
x*3
```

In contrast in a language like C, one has to declare a variable based on its type before using it:

```
double y;
double x = 3.1;
y = x * 7.1;
```

Dynamic typing can be quite helpful from the perspective of quick implementation and avoiding tedious type definitions and problems from minor inconsistencies between types (e.g., multiplying an integer by a real-valued number).
But static typing has some critical advantages from the perspective of software development, including:

  - protecting against errors from mismatched values and unexpected user inputs, and
  - generally much faster execution because the type of a variable does not need to be checked when the code is run.

More complex types in R (and in Python) often use references (*pointers*, aka *addresses*) to the actual locations of the data. We'll see this in detail later in the Unit.

### Types and classes in R

You should be familiar with vectors as the basic data structure in R,
with character, integer, numeric, etc. classes. Vectors are either
*atomic vectors* or *lists*. Atomic vectors generally contain one of the
four following types: *logical*, *integer*, *double* (i.e., *numeric*), and
*character*.

Everything in R is an object and all objects have a class. For simple
objects class and type are often closely related, but this is not the
case for more complicated objects. As we'll see later in the Unit, the class describes what the object
contains and standard functions associated with it. In general, you
mainly need to know what class an object is rather than its type.

> **Note**
> You can look at Table 7.1 in the Adler book to see some other types.

Let's look at the type and class of various data structures in R.
We'll first see that real-valued are stored as double-precision (8 byte) floating point numbers
internally in R (as 'doubles' in C, as the R interpreter is a program written in C).

```{r}
devs <- rnorm(5)
class(devs)
typeof(devs)   
a <- data.frame(x = 1:2)
class(a)
typeof(a)
is.data.frame(a)
is.matrix(a)
is(a, "matrix")
m <- matrix(1:4, nrow = 2) 
class(m) 
typeof(m)
```

In most cases integer-valued numbers are stored as numeric values in R, but there are exceptions such as the result of using the
sequence operater, `:`, above. We can force R to store values
as integers:

```{r}
vals <- c(1, 2, 3)
class(vals)
vals <- 1:3
class(vals)
vals <- c(1L, 2L, 3L)
vals
class(vals)
```


### Attributes

We saw the notion of attributes when looking at HTML and XML, where the information was stored as
key-value pairs that in many cases had additional information in the form of attributes.

In R, *attributes* are information about an object attached to an object as
something that looks like a named list. Attributes are often copied when
operating on an object. This can lead to some weird-looking formatting when 
in subsequent operations the *names* attribute is carried along:

```{r}
x <- rnorm(10 * 365)
attributes(x)
qs <- quantile(x, c(.025, .975))
attributes(qs)
qs
qs[1] + 3
object.size(qs)
```

We can get rid of the attribute:

```{r}
names(qs) <- NULL
qs
object.size(qs)
```

A common use of attributes is that rows and columns may be named in
matrices and data frames, and elements in vectors:

```{r}
df <- data.frame(x = 1:2, y = 3:4)
attributes(df)
row.names(df) <- c("first", "second")
df
attributes(df)
vec <- c(first = 7, second = 1, third = 5)
vec['first']
attributes(vec)
```

### Converting between types

This also goes by the term *coercion* and *casting*. 
Casting often needs to be done explicitly in compiled languages and somewhat less so in interpreted languages like R.

We convert between classes using variants on *as*: e.g.,

```{r}
as.character(c(1,2,3))
as.numeric(c("1", "2.73"))
as.factor(c("a", "b", "c"))
```

Some common conversions are converting numbers that are being
interpreted as characters into actual numbers, converting between
factors and characters, and converting between logical TRUE/FALSE
vectors and numeric 1/0 vectors. 

In some cases R will automatically do
conversions behind the scenes in a smart way (or occasionally not so
smart way). Consider these examples of implicit coercion:

```{r}
x <- rnorm(5)
x[3] <- 'hat' # What do you think is going to happen?
indices <- c(1, 2.73)
myVec <- 1:10
myVec[indices]
```

Here's an example we can work through that will help illustrate how type
conversions occur behind the scenes in R.

```{r}
n <- 5
df <- data.frame(label = rep('a', n), val1 = rnorm(n), val2 = rnorm(n))
df
## Why does the following not work?
try( apply(df, 1, function(x) x[2] + x[3]) )
## Instead, this will work. Why?
apply(df[ , 2:3], 1, function(x) x[1] + x[2])
```

Be careful of using factors as indices:

```{r}
students <- factor(c("basic", "proficient", "advanced",
                     "basic", "advanced", "minimal"))
score <- c(minimal = 65, basic = 75, proficient = 85, advanced = 95)
score["advanced"]
students[3]
score[students[3]]
score[as.character(students[3])]
```

What has gone wrong and how does it relate to type coercion?



## Data frames and related concepts

### Some notes on data frames and operations on data frames

Base R provides a variety of functions for manipulating data frames, but
now many researchers use add-on packages (many written by Hadley Wickham
as part of a group of packages called the *tidyverse*) to do these
manipulations in a more elegant way. [Module 6 of the R bootcamp](https://https://berkeley-scf.github.io/r-bootcamp-fall-2022/modules/module6_tidyverse) describes some of these new tools in more details, but I'll touch on some aspects of this here, without showing much of the tidyverse syntax.

### split-apply-combine

Often analyses are done in a stratified fashion - the same operation or
analysis is done on subsets of the data set. The subsets might be
different time points, different locations, different hospitals,
different people, etc.

The split-apply-combine framework is intended to operate in this kind of
context: first one splits the dataset by one or more variables, then one
does something to each subset, and then one combines the results. The
*dplyr* package implements this framework (as does the *pandas* package
for Python). One can also do similar operations using various flavors of
the *lapply* family of functions such as *by*, *tapply*, and
*aggregate*, but the dplyr-based tools are often nicer to use.

split-apply-combine is also closely related to the famous Map-Reduce framework
underlying big data tools such as Hadoop and Spark. 

It's also very similar to standard SQL queries involving filtering, grouping, and
aggregation.

### Long and wide formats

Finally, we may want to convert between so-called 'long' and 'wide'
formats, which we can motivate in the context of longitudinal data
(multiple observations per subject) and panel data (temporal data for
each of multiple units such as in econometrics). The wide format has
repeated measurements for a subject in separate columns, while the long
format has repeated measurements in separate rows, with a column for
differentiating the repeated measurements. 

```{r}
long <- data.frame(id = c(1, 1, 1, 2, 2, 2),
                   time = c(1980, 1990, 2000, 1980, 1990, 2000),
                   value = c(5, 8, 9, 7, 4, 7))
wide <- data.frame(id = c(1, 2),
                   value_1980 = c(5, 7), value_1990 = c(8, 4), value_2000 = c(9, 7))
long
wide
```

The wide format can be useful in some situations for treating each row as a (multivariate observation), 
but the long formatwhile the long format is often what is needed for analyses such as 
mixed models. ANOVA, or for plotting, such as with *ggplot2*.

There are a variety of functions for converting between wide and long
formats. I recommend *pivot_longer* and *pivot_wider* in the *tidyr* package. There are also older *tidyr* functions
called *gather* and *spread*. There are also the *melt* and
*cast* in the *reshape2* package. These are easier to use than the
functions in base R such as *reshape* or *stack* and *unstack*
functions*.*

### Piping

Piping was introduced into R in conjuction with *dplyr* and the *tidyverse*.

The tidyverse pipe is `%>%` while the new base R pipe is `|>`.
These are based on the UNIX pipe, which we saw in Unit 3, though
they behave somewhat differently in that the output of the previous function
is passed in as the *first* argument of the next function. In the shell,
the pipe connects *stdout* from the previous command to *stdin* for the next command.


### Non-standard evaluation and the tidyverse

Many tidyverse packages use non-standard evaluation to make it easier to
code. For example in the following dplyr example, you can refer directly
to *country* and *unemp*, which are variables in the data frame, without
using `data$country` or `data$unemp` and without using quotes around the
variable names, as in `"country"` or `"unemp"`. Referring directly to
the variables in the data frame is not standard R usage, hence the term
"non-standard evaluation". One reason it is not standard is that *country*
and *unemp* are not themselves independent R variables so R can't find
them in the usual way using scoping (discussed later in the Unit).

```{r}
library(dplyr)
cpds <- read.csv(file.path('..', 'data', 'cpds.csv'))

cpds2 <- cpds %>% group_by(country) %>%
                  mutate(mean_unemp = mean(unemp))

head(cpds2)
```


This 'magic' is done by capturing the code expression you write and
evaluating it in a special way in the context of the data frame. I
believe this uses R's environment class (discussed later in the Unit), but
haven't looked more deeply.

While this has benefits, this so-called non-standard evaluation makes it
harder to program functions in the usual way, as illustrated in the
following code chunk, where neither attempt to use the function works.

```{r}
add_mean <- function(data, group_var, summarize_var) {
    data %>% group_by(group_var) %>%
             mutate(mean_of_var = mean(summarize_var))
}

try(cpds2 <- add_mean(cpds, country, unemp))
try(cpds2 <- add_mean(cpds, 'country', 'unemp'))
```


For more details on how to avoid this problem when writing functions
that involve tidyverse manipulations, see
[this tidyverse programming guide](https://dplyr.tidyverse.org/articles/programming.html).

Note that the tidyverse is not the only place where non-standard
evaluation is used. Consider this *lm* call:

```{r, eval=FALSE}
lm(y ~ x, weights = w, data = mydf)
```

> **Challenge**
> Where is the non-standard evaluation there?

# 5. Programming paradigms: object-oriented and functional programming

Object-oriented and functional programming are two important approaches to programming.

Functional programming focuses on writing functions that take inputs and produce outputs. Ideally those functions don't change the state (i.e., the values) of any variables and can be treated as black boxes. Functions can be treated like other variables, such as passing functions as arguments (as one does with *lapply* in R, for example).

Object-oriented programming revolves around objects that belong to classes. The class of an object defines the fields (the data objects) holding information and (often) methods that can be applied to those fields. When one calls a method, it may modify the value of the fields. A statistical analogy is that an object of a class is like the realization (the object) of a random variable (the class).

One can think of functional programming as being focused on actions (or *verbs* to make an analogy with human language). One carries out a computation as a sequence of function calls. One can think of OOP as being focused on the objects (or *nouns*). One carries out a computation as a sequence of operations with the objects, using the class methods.

Many languages are multi-paradigm, containing aspects of both approaches and allowing programmers to use either approach. Both R and Python are like this, though some might consider R to be more functional and Python to be more object-oriented. That said, in R everything is an object and has a class, while there are plenty of function-based operations in Python.

```python
import numpy as np
x = np.array([1.2, 3.5, 4.2])
x.shape  # field (or attribute) of the numpy array class
x.sum()  # method of the class
len(x)   # function
```

Different people have different preferences, but which is better depends on what you are trying to do. If your computation is a data analysis pipeline that involves a series of transformations of some data, a functional approach might make more sense, since the focus is on a series of actions rather than the state of objects. If your computation involves various operations on fixed objects whose state needs to change, OOP might make more sense. For example, if you were writing code to keep track of student information, it would probably make sense to have each student as an object of a Student class with methods such as 'register' and 'assign_grade'. 


# 6. Object-oriented programming (OOP)

## Principles

Some of the standard concepts in object-oriented programming include *encapsulation*, *inheritance*, *polymorphism*, and *abstraction*.

*Encapsulation* involves preventing direct access to internal data in an object from outside the object. Instead the class is designed so that access (reading or writing) happens through the interface set up by the programmer (e.g., 'getter' and 'setter' methods). We'll see this in our R6 class example below.

*Inheritance* allows one class to be based on another class, adding more specialized features. An example in R's S3 system is that the *glm* class inherits from the *lm* class. 

*Polymorphism* allows for different behavior of an object or function depending on the context. A polymorphic function behaves differently depending on the input types. A polymorphic object is one that can belong to different classes (e.g., based on inheritance), and a given method name can be used with any of the classes. An example would be having a base or super class called 'algorithm' and various specific machine learning algorithms inheriting from that class. All of the classes might have a 'predict' method.

*Abstraction* involves hiding the details of how something is done (e.g., via the method of a class), giving the user an interface to provide inputs and get outputs. By making the actual computation a black box, the programmer can modify the internals without changing how a user uses the system. 

Classes generally have *constructors* that initialize objects of the class and *destructors* that remove objects.


## Generic function OOP

Much of the object-oriented programming in R uses *generic function OOP*, also known as *functional OOP*. 
In this style, classes don't have methods. Instead there are *generic functions* (also known as *generic methods*) that change their behavior based on the type of the input(s). Another way to put it is that the nouns and the verbs are separate, unliked in standard OOP.

The use of generic functions is similar in spirit to function or method *overloading* in C++ and Java.

Generic function OOP  is how the (very) old S3 system in R works. It's also a key part of the (fairly) new Julia language.

### S3 classes in R

S3 classes are widely-used, in particular for statistical models in the
*stats* package. S3 classes are very informal in that there's not a
formal definition for an S3 class. Instead, an S3 object is just a
primitive R object such as a list or vector with additional attributes
including a class name.

#### Creating our own class

We can create an object with a new class as follows:

```{r}
yog <- list(firstname = 'Yogi', surname = 'the Bear', age = 20)
class(yog) <- 'bear' 
```

Actually, if we want to create a new class that we'll use again, we want
to create a *constructor* function that initializes new bears:

```{r}
bear <- function(firstname = NA, surname = NA, age = NA){
	# constructor for 'indiv' class
	obj <- list(firstname = firstname, surname = surname,
                    age = age)
	class(obj) <- 'bear' 
	return(obj)
}
smoke <- bear('Smokey','Bear')
```

For those of you used to more formal OOP, the following is probably
disconcerting:

```{r}
class(smoke) <- "celebrity"
```

Generally S3 classes inherit from lists (i.e., are special cases of lists),
so you can obtain components of the object using the \$ operator.

#### Generic methods

The real power of the S3 system comes from defining *class-specific methods*. For example,

```{r}
x <- rnorm(10)
summary(x)

y <- rnorm(10)
mod <- lm(y ~ x)
summary(mod)
```

Here *summary* is a generic function (or generic method) that, based
on the type of object given to it (the first argument), dispatches a
class-specific function (method) that operates on the object. 

The above is equivalent to directly calling the class-specific methods:

```{r}
identical(summary(x), summary.default(x))
identical(summary(mod), summary.lm(mod))
```


This use of generic functions is convenient in that it allows us to work with a variety of kinds of objects using familiar functions. Consider
the generic methods *plot*, *print*, *summary*, *\[*, and
others. We can look at a function and easily see that it is a generic
method. 

```{r}
summary
```

The `UseMethod` syntax is what causes the dispatching of the class-specific method
associated with `object` and calls that method.
In many cases there will be a default method (here,
*summary.default*), so if no method is defined for the class, R uses
the default. Sidenote: arguments to a generic method are passed along to
the selected method by passing along the calling environment.

We can also see what classes have methods for a given generic
function.

```{r}
methods(summary)
```

Or from a different angle we can see what specific methods are available for a given class.

```{r}
methods(class = 'lm')
```

Let's try this functionality  out on our *bear* class.

```{r}
summary.bear <- function(object) 
	return(with(object, cat("Bear of age ", age, 
	" whose name is ", firstname, " ", surname, ".\n",
    sep = "")))
summary(yog)
```

We can also define a new generic function.

Let's do this for the *bear* class as an illustration, though this won't
provide any functionality beyond what we did with *summary*

```{r}
summarize <- function(object, ...) 
	UseMethod("summarize") 
```

```{r}
summarize.bear <- function(object) 
	return(with(object, cat("Bear of age ", age, 
	" whose name is ", firstname, " ", surname, ".\n",
    sep = "")))
summarize(yog)
```

#### Why use generic functions?

We could have written *summary* as a regular function with a bunch of if statements or if-else clauses (or *switch*) so that it can handle different kinds of input objects. 

This has two disadvantages:

  1. We need to write the code that does the checking (and all the code for the different cases all lives inside one potentially very long function, unless we create class-specific helper functions).
  2. Much more importantly, *summary* will only work for existing classes. And users can't easily extend it for new classes that they create because they don't control the *summary* function. So a user could not add the additional conditions/classes in a big if-else statement. The generic function approach makes the system *extensible* -- we can build our own new functionality on what is already in R. For example, we could have written *summary.bear*.

#### The print method

Like *summary*, *print* is a generic method, with various
class-specific methods, such as *print.lm*. We could write our own
*print.bear* specific method. 

Note that the *print* function is what is called when you simply type
the name of the object, so we can have object information printed out in
a structured way. Thus, the output when we type the name of an
*lm* object is NOT simply a regurgitation of the elements of the list -
rather *print.lm* is called.

```{r}
mod
print(mod)
stats:::print.lm(mod)  ## print.lm is private to the stats namespace
# print.default(mod)   ## lots of output, so don't print in document...
```

```{r}
stats:::print.lm
```

Surprisingly, the *summary* method generally doesn't actually print
out information; rather it computes things not stored in the original
object and returns it as a new class (e.g., class *summary.lm*), which
is then automatically printed, per my comment above (e.g., using
*print.summary.lm*), unless one assigns it to a new object. Note that
*print.summary.lm* is hidden from user view (it's a private object in the *stats* namespace).

```{r}
out <- summary(mod)
class(out)
out
print(out)
## One can look at the code for the method (not shown):
## getS3method(f = "print", class = "summary.lm")
```

#### Inheritance

Let's look at the *lm* class, which builds on lists, and *glm* class,
which builds on the *lm* class. Here *mod* is an object (an instance) of
class *lm*. 

```{r}
library(methods)
ybin <- sample(c(0, 1), 10, replace = TRUE)
ycont <- rnorm(10)
x <- rnorm(10)
mod1 <- lm(ycont ~ x)
mod2 <- glm(ybin ~ x, family = binomial)
class(mod1)
class(mod2)
is.list(mod1)
names(mod1)
is(mod2, "lm")
```

Here's an example of why this is useful. We don't have to define methods for
the *glm* class if the given method for the *lm* class would work fine:

```{r}
model.matrix(mod1)
model.matrix(mod2)
methods(model.matrix)
```

As noted with *lm* and *glm* objects, we can assign more than one class
to an object. Here *summarize* still works, even though the primary
class is *grizzly_bear*.

```{r}
class(yog) <- c('grizzly_bear', 'bear')
summarize(yog) 
```

The classes should nest within one another with the more specific
classes to the left, e.g., here a *grizzly_bear* would have some
additional fields on top of those of a *bear*, perhaps
*number_of_people_killed* (since grizzly bears are much more dangerous
than some other kinds of bears), and perhaps additional or modified
methods. *grizzly_bear* inherits from *bear*, and R uses methods for the
first class before methods for the next class(es).

The above is an example of polymorphism. `yog` is a polymorphic object 
and the various methods are polymorphic in that *print* can be used
with the *bear* class, the *grizzly_bear* class, and other classes beyond that.


> **Challenge**
> How would you get R to quit immediately, without asking for any more information, when you simply type `k` (no parentheses!) instead of `quit()`? (Hint: you can do this by understanding what happens when you type `k` and how to exploit the S3 system.)


### Multiple dispatch OOP 

S3 method dispatch involves only the first argument to the function. In contrast, [Julia emphasizes the importance of multiple dispatch](https://docs.julialang.org/en/v1/manual/methods) as particularly important for mathematical computation. With multiple dispatch, the specific method can be chosen based on more than one argument.

The old (but still used in some contexts) [S4](http://adv-r.had.co.nz/S4.html) system in R and the (very) new [R7](https://rconsortium.github.io/OOP-WG) system both provide for multiple dispatch. 

As a very simple example unrelated to any specific language, multiple dispatch would allow one to do the following with the addition operator:

```
3 + 7    # 10
3 + 'a'  # '3a'
'hi' +  ' there'  # 'hi there'
```

The idea of having the behavior of an operator or function adapt to the type of the input(s) is one aspect of *polymorphism*.

Both S4 and R7 are designed to be more formal than the S3 system (recall how we could just 'create' an S3 class by giving a class name to an existing list). With S4 and R7, you need to define your classes. 

## 'Standard' OOP

What I'm calling 'standard' object-oriented programming is the style of OOP used in languages such as Python, C++, and Java. In R, one can use this style via the R6 system (or the older *referenceClass* system).

In this style, objects belong to classes. A class is made up of fields (the data objects) that store information and methods that operate on the fields. Thus, unlike generic function OOP, the verbs are part of the nouns.

We'll illustrate this style of OOP using an example with an R6 class.

### R6 classes

R6 classes are a somewhat new construct in R, with a class-based approach fairly similar to Python and C++. Importantly, they behave like pointers. We'll discuss pointers in detail later. Let's work through an example
where we set up the fields of the class and class
methods, including a constructor.

#### Example

Our example is to create a class for working with random time series. Each object of the class has specific parameter values that control the stochastic behavior of the time series. With a given object we can simulate one or more time series (realizations).

Here's the initial definition of the class, with both public
(user-facing) and private (internal use only) methods and fields.

```{r, include=FALSE}
# suppress package-loading messages by loading here
library(fields)  
```

```{r}
library(R6)

tsSimClass <- R6Class("tsSimClass",
    ## class for holding time series simulators
    public = list(
        initialize = function(times, mean = 0, corParam = 1) {
            library(fields)
            stopifnot(is.numeric(corParam), length(corParam) == 1)
            stopifnot(is.numeric(times))
            private$times <- times
            private$n <- length(times)
            private$mean <- mean
            private$corParam <- corParam
            private$currentU <- FALSE
            private$calcMats()
        },
        
        changeTimes = function(newTimes) {
            private$times <- newTimes
            private$calcMats()
        },
        
        getTimes = function() {
            return(private$times)
        },

        print = function() { # 'print' method
            cat("R6 Object of class 'tsSimClass' with ",
                private$n, " time points.\n", sep = '')
            invisible(self)
        },

        simulate = function() {
            if(!private$currentU)     
                private$calcMats()
            ## analogous to mu+sigma*z for generating N(mu, sigma^2)
            return(private$mean + crossprod(private$U, rnorm(private$n)))
        }
    ),

    ## private methods and functions not accessible externally
    private = list(
        calcMats = function() {
            ## calculates correlation matrix and Cholesky factor
            lagMat <- fields::rdist(private$times) # local variable
            corMat <- exp(-lagMat^2 / private$corParam^2)
            private$U <- chol(corMat) # square root matrix
            cat("Done updating correlation matrix and Cholesky factor.\n")
            private$currentU <- TRUE
            invisible(self)
        },
        n = NULL, 
        times = NULL,
        mean = NULL,
        corParam = NULL,
        U = NULL,
        currentU = FALSE
    )
)   

```


Now let's see how we would use the class.

```{r}
myts <- tsSimClass$new(1:100, 2, 1)
myts
set.seed(1)
## here's a simulated time series
y1 <- myts$simulate()
plot(myts$getTimes(), y1, type = 'l', xlab = 'time',
     ylab = 'process values')
## simulate a second series
y2 <- myts$simulate()
lines(myts$getTimes(), y2, lty = 2)
```

We could set up a different object that has different parameter values.
That new simulated time series is less wiggly because the `corParam` value
 is larger than before.


```{r}
myts2 <- tsSimClass$new(1:100, 2, 4)
set.seed(1)
## here's a simulated time series with a different value of
## the correlation parameter (corParam)
y3 <- myts2$simulate()

plot(myts$getTimes(), y1, type = 'l', xlab = 'time',
     ylab = 'process values')
lines(myts$getTimes(), y2, lty = 2)
lines(myts2$getTimes(), y3, col = 'red')
```

#### Copies and references

Next let's think about when copies are made. In the next example `mytsRef` is a copy of `myts`
in the sense that both names point to the same underlying object.
But no data were copied when the assignment to `mytsRef` was done. 

```{r}
mytsRef <- myts
## 'mytsRef' and 'myts' are names for the same underlying object
mytsFullCopy <- myts$clone()  

## Now let's change the values of a field
myts$changeTimes(seq(0,1000, length = 100))
myts$getTimes()[1:5] 
mytsRef$getTimes()[1:5] # the same as `myts`
mytsFullCopy$getTimes()[1:5] # different from `myts`
```

In contrast `mytsFullCopy` is a reference to a different object, and
all the data from `myts` had to be copied over to `mytsFullCopy`. This takes additional memory (and time), but is also safer, as it avoids the possibility that the user might modify `myts` and not realize that they were also affecting `mytsRef`.

#### Encapsulation

Why have private fields (i.e., encapsulation)? The use of private fields shields them from
    modification by users. In this case, that prevent users from modifying the
*times* field. Why is this important? In this example, the correlation matrix and
    the Cholesky factor U are both functions of the vector of times. So
    we don't want to allow a user to directly modify *times*. If they did, it would leave the fields of the object in inconsistent states. Instead we
    force them to use *setTimes*, which correctly keeps all the fields
    in the object internally consistent (by calling *calcMats*). It also allows us to improve efficiency
by controlling when computationally expensive operations are carried out.

```{r}
try(myts$times <- 1:10)
```

#### Final comments

  -   As we saw above, a copy of an object is just a pointer to the original object, unless we explicitly invoke the *clone* method.
  -   Classes can inherit from other classes. E.g., if we had a *simClass* and we wanted the *tsSimClass* to inherit from it:

      ```
      R6Class(tsSimClass, inherit = simClass, ...) 
      ```
  -   If you need to refer to methods and fields you refer to the entire object as either *self* or *private*.


More details on R6 classes can be found in the [Advanced R book](https://adv-r.hadley.nz/r6.html).


# 7. Functional programming

## Overview of functional programming

Functional programming is an approach to programming that emphasizes the use of modular, self-contained functions.
Such functions should operate only on arguments provided to
them (avoiding global variables), and **produce no side effects**, although in some cases there are good
reasons for making an exception. Another aspect of functional programming is 
that functions are considered 'first-class' citizens in that they can be passed as arguments to another function,
returned as the result of a function, and assigned to variables. In other words, a function can be treated as any other variable.

In many cases (including R and Python), anonymous functions (also called 'lambda functions') can be created on-the-fly for use in various circumstances. 

## Functional programming in R

R is a language that has strong functional programming aspects to it, including:

  - All operations are carried out by functions.
  - Functions are first class citizens.
  - Functions (generally) do not have side effects.

Functions that are not implemented internally in R  are also referred to officially as *closures* (this is their
*type*) - this terminology sometimes comes up in error messages.

```{r}
typeof(mean)
typeof(lm)
typeof(length)
```


### No side effects

Most functions available in R (and ideally functions that you write as well) operate by taking in arguments and producing output that is then (presumably) used subsequently. The functions generally don't have any effect on the state of your R environment/session other than the output they produce.

An important reason for this (plus for not using global variables) is that it means that it is easy for people using the language to understand what code does. Every function can be treated a black box -- you don't need to understand what happens in the function or worry that the function might do something unexpected (such as changing the value of one of your variables). The result of running code is simply the result of a composition of functions, as in mathematical function composition.

One aspect of this is that R uses a *pass-by-value* approach to function arguments (as opposed to a *pass-by-reference* approach). We'll talk about function arguments and when copies are made  in much more detail later, but briefly, when you pass an object in as an argument and then modify it in the function, you are modifying a local copy of the variable that exists in the context (the *frame*) of the function and is deleted when the function call finishes:

```{r}
x <- 1:3
myfun <- function(x) {
      x[2] <- 7
      print(x)
      return(x)
}

new_x <- myfun(x)
x   # unmodified
```

In contrast, let's see what happens in Python

```{python}
x = [1,2,3]
def myfun(x):
  x[1] = 7
  print(x)
  return(x)

new_x = myfun(x)
x   # modified!
```

There are some (necessary) exceptions to the idea of no side effects in R. 
An important exception is *par()*. If you change graphics parameters by
calling *par()* in a user-defined function, they are changed permanently
outside of the function. One trick is as follows:

```{r}
f <- function(){
  oldpar <- par()
  par(cex = 2)
          
  # body of code

  par() <- oldpar
}
```
Note that changing graphics parameters within a specific plotting
function - e.g., `plot(x, y, pch = '+')`, doesn't change things except
for that particular plot.

> *Challenge*
> What are some other functions that are called for the purpose of the side effects they produce? (For example, which functions change the state of your R session in some way?

### Functions are first-class objects

Everything in R is an object, including functions. We can assign
functions to variables in the same way we assign numeric and other
values.

```{r}
x <- 3
class(x); typeof(x)

try(x(2))    # x is not a function (yet)
x <- function(z) z^2  # now it is a function
x(2)
class(x); typeof(x)
```

We can call a function based on the text name of the function.

```{r}
myFun <- 'mean'; x <- rnorm(10)
eval(as.name(myFun))(x)
```

We can also pass a function into another function as the actual function
object. This is an important aspect of R being a functional programming language.

```{r}
x <- rnorm(10)
sapply(x, abs)

f <- function(fxn, x) {
    fxn(x)
}
f(mean, x)
```

We can also pass in a function based on a a character vector of length
one with the name of the function. Here *match.fun()* is a handy
function that extracts a function when the function is passed in as an
argument of a function. It looks in the calling environment for the
function and can handle when the function is passed in as a function
object or as a character vector of length 1 giving the function name.



```{r}
f <- function(fxn, x){
  match.fun(fxn)(x) 
}
f("mean", x)
f(mean, x)
```


Function objects contain three components: an argument list, a body (a
parsed R statement), and an environment.

```{r}
f1 <- function(x) y <- x^2
f2 <- function(x) {
    y <- x^2
    z <- x^3
    return(list(y, z))
}
class(f1)
body(f2)
typeof(body(f1)); class(body(f1))
typeof(body(f2)); class(body(f2))
```

We'll see more about objects relating to the R language and parsed code
in the final section of this Unit. For now, just realize that the parsed
code itself is treated as an object(s) with certain types and certain
classes.


The *do.call* function is another example of a function that takes a function as an argument.
It will apply a function to the elements of a
list. For example, we can `rbind()` together (if compatible) the
elements of a list of vectors instead of having to loop over the
elements or manually type them in:

```{r}
myList <- list(a = 1:3, b = 11:13, c = 21:23)
args(rbind)
rbind(myList$a, myList$b, myList$c)
rbind(myList)
do.call(rbind, myList)
```

Why couldn't we just use *rbind* directly? Basically we're using
`do.call()` to use functions that take `...` as input (i.e., functions
accepting an arbitrary number of arguments) and to use the list as the
input instead (i.e., to use the list elements).

More generally *do.call* is a way to pass arguments to a function when 
the arguments you want to pass are part of a list.

```{r}
do.call(mean, list(1:10, na.rm = TRUE))
```


### All operations are functions

All operations in R are actually function calls, even things that don't look like function calls,
including various operators (such as addition,
subtraction, etc.), printing to the screen, etc.

#### Operators

Operators, such as `+` and `[` are just functions, but their arguments
can occur both before and after the function call:

```{r}
a <- 7; b <- 3
# let's think about the following as a mathematical function
#  -- what's the function call?
a + b 
`+`(a, b)
```

In general, you can use back-ticks to refer to the operators as
operators instead of characters. In some cases single or double quotes
also work. We can look at the code of an operator as follows using
back-ticks to escape out of the standard R parsing, e.g., ``%*%``.

Finally, since an operator is just a function, you can use it as an
argument in various places:

```{r}
x <- 1:3; y <- c(100,200,300)
outer(x, y, `+`)

myList <- list(list(state = 'new york', value = 1:5),
               list(state = 'california', value = 6:10),
               list(state = 'delaware', value = 11:15))

result <- lapply(myList, `[[`, 2)
result
## note that the index "2" is the additional argument to the [[ function
myMat <- sapply(myList, `[[`, 2)
myMat
cbind(myList[[1]][[2]], myList[[2]][[2]])  ## equivalent but doesn't scale
```

You can define your own *binary* operator (an operator taking two
arguments) using a string inside *%* symbols. Here's how we could do
Python-style string addition:

```{r}
`%+%` <- function(a, b) paste0(a, b, collapse = '')
"Hi " %+% "there"
```

Since operators are just functions, there are cases in which there are
optional arguments that we might not expect. Here's how to pass a
sometimes useful argument to the bracket operator (in this case avoiding
conversion from a matrix to a vector, which can mess up subsequent
code).

```{r}
mat <- matrix(1:4, 2, 2)
mat[ , 1] 
mat[ , 1, drop = FALSE] # what's the difference?
```

We can also use operators with our S3 classes. Picking up our example
from our discussion of S3 OOP, the following example will be a bit silly
(it would make more sense with a class that is a mathematical object)
but indicates the power of having methods.

```{r}
yog <- list(firstname = 'Yogi', surname = 'the Bear', age = 20)
class(yog) <- 'bear' 

methods(`+`)
`+.bear` <- function(object, incr) {
	object$age <- object$age + incr
	return(object)
}
older_yog <- yog + 15

older_yog
```

#### Other operations that are functions

Even beyond operators, all code in R can be viewed as a function call, including
if statements and for and while loops.

What do you think is the functional version of the following code? What
are the arguments?

```{r, eval=FALSE}
if(x > 27){
	print(x)	
} else{
	print("too small") 
}
```

#### Replacement functions

Assignments that involve functions or operators on the left-hand side
(LHS) are called *replacement expressions* or *replacement functions.*
These can be quite handy. Here are a few examples:

```{r}
diag(mat) <- c(3, 2)
is.na(vec) <- 3
names(df) <- c('var1', 'var2')
```

Replacement expressions are actually function calls. The R interpreter
calls the replacement function (which often creates a new object that
includes the replacement) and then assigns the result to the name of the
original object.

```{r}
mat <- matrix(rnorm(4), 2, 2)
diag(mat) <- c(3, 2)
mat
mat <- `diag<-`(mat, c(10, 21))
mat
base::`diag<-`
```

The old version of *mat* still exists until R's memory management cleans
it up, but it's no longer referred to by the symbol *mat*.
This can cause memory use to increase temporarily (but generally very briefly).
So it's something to keep in mind if you're doing replacements on large objects.

You can define your own replacement functions like this, with the
requirements that the last argument be named '*value*' and that the
function return the entire object:

```{r}
yog <- list(firstName = 'Yogi', lastName = 'Bear')
`firstName<-` <- function(obj, value){
  obj$firstName <- value
  return(obj)
}
firstName(yog) <- 'Yogisandra'
```

We can use replacement functions with S3 classes we define. Again,
picking up our example from our discussion of S3 OOP, this is again a
bit silly but we could do the following. We need to define the generic
replacement function and then the class-specific one.

```{r}
`age<-` <- function(x, ...) UseMethod("age<-")

`age<-.bear` <- function(object, value){ 
	object$age <- value
	return(object)
}
age(older_yog) <- 60

older_yog
```


## Function evaluation, frames, and the call stack

### Overview

When we run code, we end up calling functions inside of other function calls.
This leads to a nested series of function calls. The series of calls is the *call stack*.
The stack operates like a stack of cafeteria trays - when a
function is called, it is added to the stack (pushed) and when it
finishes, it is removed (popped).

Understanding the series of calls is important when reading error messages and debugging.
In Python, when an error occurs, the call stack is shown, which has the advantage of
giving the complete history of what led to the error and the disadvantage of producing
often very verbose output that can be hard to understand. In R, only the function in which
the error occurs is shown, but you can see the full call stack by invoking `traceback()` (see the [debugging tutorial](https://github.com/berkeley-scf/tutorial-R-debugging)).

What happens when an R function is evaluated? 

  - The user-provided function arguments are evaluated in the calling environment and the results are
matched to the argument names in the function definition. 
  - A new environment with its own frame is created, with the frame on the call stack. Assignment to the argument names is done in the environment, including any default arguments. 
  - The body of the function is evaluated in the environment. Any look-up of variables not found in the
environment is done using R's lexical scoping rules to look in the
series of enclosing environments. 
  - When the function finishes, the return
value is passed back to the calling frame and the function frame is
taken off the stack. The environment is removed, unless the environment
serves as the enclosing environment of another environment.

I'm not expecting you to fully understand that previous paragraph and
all the terms in it yet. We'll see all the details as we proceed through this Unit.

### Frames and the call stack

R keeps track of the call stack. Each function call is associated with
a *frame* that contains the local variables for that function call.

There are a bunch of functions that let us query what frames are on the
stack and access objects in particular frames of interest. This gives us
the ability to work with objects in the frame from which a function was
called.

Some terminology: for our purposes we'll use the terms *frame* and
*environment* somewhat interchangeably for the moment. A *frame* or
*environment* is a collection of named objects. (Note that when we talk
about variable scope later in this Unit, we'll have to be more careful with
our terminology.) So in the context of a function call, the frame is the
set of local variables available in the function, including arguments
passed to the function.

R provides some functions that allow you to query the call stack and its frames.
*sys.nframe* returns the number of the current frame/environment and
*sys.parent* the number of the parent, while *parent.frame* gives
the name of the frame/environment of the parent (i.e., the calling)
frame. *sys.frame* gives the name of the frame/environment for a given
frame number (for non-negative numbers). For negative numbers, it goes
back that many frames in the call stack and returns the name of the
frame/environment. I need to manually insert the output here because the R Markdown processing
up the frame counting somehow.


```{r, eval=FALSE}
sys.nframe()
f <- function() {
	cat('in f: Frame number is ', sys.nframe(),
            '; parent frame number is ', sys.parent(), '.\n', sep = '')
	cat('in f: Frame (i.e., environment) is: ')
	print(sys.frame(sys.nframe()))
	cat('in f: Parent is ')
	print(parent.frame())
	cat('in f: Two frames up is ')
	print(sys.frame(-2))
}
f()
```

```
in f: Frame number is 1; parent frame number is 0.
in f: Frame (i.e., environment) is: <environment: 0x55a4d71beb88>
in f: Parent is <environment: R_GlobalEnv>
in f: Two frames up is Error in sys.frame(-2) : not that many frames on the stack
```

```{r, eval=FALSE}
ff <- function() {
	cat('in ff: Frame (i.e., environment) is: ')
	print(sys.frame(sys.nframe()))
	cat('in ff: Parent is ')
	print(parent.frame())	
	f()
}
ff() 
```

```
in ff: Frame (i.e., environment) is: <environment: 0x55a4d7391700>
in ff: Parent is <environment: R_GlobalEnv>
in f: Frame number is 2; parent frame number is 1.
in f: Frame (i.e., environment) is: <environment: 0x55a4d7393b38>
in f: Parent is <environment: 0x55a4d7391700>
in f: Two frames up is <environment: R_GlobalEnv>
```

Next we'll use a recursive function to illustrate what information we can gather
about the call stack using *sys.status*. *sys.status* gives extensive information about the call
stack and the frames involved (*sys.status* uses *sys.calls*,
*sys.parents* and *sys.frames*).


```{r, eval=FALSE}
g <- function(y) {
    if(y > 0) g(y-1) else gg()
}

## Ultimately, gg() is called, and it prints out info about the call stack
gg <- function() {
    ## this gives us the information from sys.calls(),
    ##   sys.parents() and sys.frames() as one object
    ## Rather than running print(sys.status()),
    ## which would involve adding print() to the call stack,
    ## we'll run sys.status and then print the result out.
    tmp <- sys.status()
    print(tmp)
}

g(3)
```

```
$sys.calls
$sys.calls[[1]]
g(3)

$sys.calls[[2]]
if(y > 0) g(y-1) else gg()

$sys.calls[[3]]
if(y > 0) g(y-1) else gg()

$sys.calls[[4]]
if(y > 0) g(y-1) else gg()

$sys.calls[[5]]
if(y > 0) g(y-1) else gg()

$sys.calls[[6]]
tmp <- sys.status()


$sys.parents
[1] 0 1 2 3 4 5

$sys.frames
$sys.frames[[1]]
<environment: 0x55a4d63479e8>

$sys.frames[[2]]
<environment: 0x55a4d6347ba8>

$sys.frames[[3]]
<environment: 0x55a4d5736638>

$sys.frames[[4]]
<environment: 0x55a4d5732028>

$sys.frames[[5]]
<environment: 0x55a4d5732098>

$sys.frames[[6]]
<environment: 0x55a4d5732488>
```

> *Challenge*
> Why did I not do `print(sys.status())` directly?

If you're interested in parsing a somewhat complicated example of frames
in action, Adler provides a user-defined timing function that evaluates
statements in the calling frame.

## Function inputs and outputs

Under construction.

## Variable scope and lookup

Under construction.

# 8. Memory and copies

Under construction.

# 9. Efficiency

Under construction.

# 10. Computing on the language (optional)

Under construction.
